Unshuffling Permutations

### 2016-02-02 submission.tex

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words naturally leads to the following computational
problems:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of time complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
% \begin{equation}
%     O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
% \end{equation}
$O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)$
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
% \begin{equation}
%     O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
% \end{equation}
$O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)$
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

@@

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator~$\Delta$, allowing to
{\em unshuffle} permutations. This operator is in fact a coproduct,
endowing the linear span of all permutations with a coalgebra structure
(see~\cite{Joni:Rota:1979} or~\cite{Grinberg:Reiner:2014} for the
definition of these algebraic structures). By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest from a combinatorial point of view.
Section~\ref{section:Binary square words and permutations} is devoted to
show that the problems related to the shuffle of words has links with
the shuffle of permutations. In particular, we show that binary words
that are square are in one-to-one correspondence with square
permutations avoiding some patterns
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. We show that $\SHUFFLE$ is
associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that if a
permutation is a square, its mirror, complement, and inverse are also
squares (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents the most important result
of this paper: the fact that deciding if a permutation is a square is
\NPC (Proposition~\ref{proposition:hardness}). This result is obtained
by exhibiting a reduction from the pattern involvement
problem~\cite{Bose:Buss:Lubiw:1998} which is \NPC.

@@

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$ are
specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the $i$-th
letter of any permutation $\pi$. For any nonnegative integer $n$, we
write $\nearrow_{n}$ (resp. $\searrow_{n}$) for the permutation
$1 2 \dots n$ (resp. $n (n-1) \dots 1$). If $\pi$ is a permutation of
$S_n$, we denote by $\bar \pi$ the {\em complement} of $\pi$, that is
the permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

@@

The number of square binary words is Sequence A191755 of~\cite{Slo}
beginning by
\begin{equation}
1, 0, 2, 0, 6, 0, 22, 0, 82, 0, 320, 0, 1268, 0, 5102, 0, 020632.
\end{equation}
By Proposition~\ref{prop:bijection_binary_to_permutations_squares}, this
is also the sequence enumerating square permutations avoiding $213$
and $231$.

@@

  It is immediate that $\mu$ can be constructed in polynomial-time in $n$ and $k$.
  % We claim that $\sigma$ occurs in $\pi$ if and only if
  % there exists a containment-free perfect matching
  % $\mathcal{M}$ in $K_\mu$ such that
  % for any two disctinct edges
  % $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  % we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  It can be shown (see Appendix) that
  $\sigma$ occurs in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $K_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \qed
\end{proof}

@@

There are a number of further directions of investigation in this
general subject. They cover several areas: algorithmic, combinatorics,
and algebra. Let us mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?
Moreover, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.

>>>>

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a well-known operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words naturally leads to the following computational
problems:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they radically differ in terms
of time complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
% \begin{equation}
%     O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
% \end{equation}
$O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)$
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
% \begin{equation}
%     O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
% \end{equation}
$O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)$
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

@@

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator~$\Delta$, allowing to
{\em unshuffle} permutations. This operator is in fact a coproduct,
endowing the linear span of all permutations with a coalgebra structure
(see~\cite{Joni:Rota:1979} or~\cite{Grinberg:Reiner:2014} for the
definition of these algebraic structures). By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest from a combinatorial point of view.
Section~\ref{section:Binary square words and permutations} is devoted to
showing that the problems related to the shuffle of words has links with
the shuffle of permutations. In particular, we show that binary words
that are square are in one-to-one correspondence with square
permutations avoiding some patterns
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. We show that $\SHUFFLE$ is
associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that if a
permutation is a square, its mirror, complement, and inverse are also
squares (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents the most important result
of this paper: the fact that deciding if a permutation is a square is
\NPC (Proposition~\ref{proposition:hardness}). This result is obtained
by exhibiting a reduction from the pattern involvement
problem~\cite{Bose:Buss:Lubiw:1998} which is \NPC.

@@

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$ are
specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the $i$-th
letter of any permutation $\pi$. For any nonnegative integer $n$, we
write $\nearrow_{n}$ (resp. $\searrow_{n}$) for the permutation
$1 2 \dots n$ (resp. $n\,(n-1) \dots 1$). If $\pi$ is a permutation of
$S_n$, we denote by $\bar \pi$ the {\em complement} of $\pi$, that is
the permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

@@

The number of square binary words is Sequence A191755 of~\cite{Slo}
beginning by
\begin{equation}
1, 0, 2, 0, 6, 0, 22, 0, 82, 0, 320, 0, 1268, 0, 5102, 0, 020632.
\end{equation}
According to
Proposition~\ref{prop:bijection_binary_to_permutations_squares}, this
is also the sequence enumerating square permutations avoiding $213$
and $231$.

@@

  It is immediate that $\mu$ can be constructed in polynomial-time in $n$ and $k$.
  % We claim that $\sigma$ occurs in $\pi$ if and only if
  % there exists a containment-free perfect matching
  % $\mathcal{M}$ in $K_\mu$ such that
  % for any two disctinct edges
  % $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  % we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  It can be shown
  %(see Appendix)
  that $\sigma$ occurs in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $K_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \qed
\end{proof}

@@

There are a number of further directions of investigation in this
general subject. They cover several areas: algorithmic, combinatorics,
and algebra. Let us mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; see also Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?
As for algebra, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.
